---
title: "eds213_data_cleaning_assign_SOFIARODAS"
format: html
author: Sofia Rodas
---
# Data Cleaning 
## Cleaning the cover columns

Load libraries and data.
```{r}
#| message: false  

# load libraries
library(tidyverse)

# create a file path
datadir_processed <- file.path("data", "processed/")

# create a variable with file names
snowsurvey_file <- "snow_cover.csv"

# import the files
snowsurvey_csv <- read_csv(file.path(datadir_processed, snowsurvey_file))
```

#### View the dataframe
Before beginning to change any data types let's inspect the data frame.
```{r}
spec(snowsurvey_csv)
```
Only `Year` and `Snow_cover` are currently numeric. The remainder of the columns are characters.


## 1. Clean the `Water_cover` column to transform it into the correct data type and respect expectations for a percentage.

Let's inspect the `Water_cover` column more closely.
```{r}
# look at all of the unique values stored in the water_cover column
unique(snowsurvey_csv$Water_cover)
```
There are many non-numeric entries in this data frame (unk, n/a, ., -). We need to make them all `NA`.


```{r}
# change all non-numeric entries to NA
snowsurvey_fixed <- snowsurvey_csv |> 
  mutate(Water_cover = ifelse(Water_cover == ".", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "unk", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover))
```

Let's make sure that no more non-numeric values other than NA are found in this column. 
```{r}
# view all non-numeric values
snowsurvey_fixed  |> 
  count(Water_cover) |> 
  filter(is.na(as.numeric(Water_cover)))
```
Amazing! All of our data is NA when it's not numeric. 

### Mutate `Water_cover` to be numeric

Now that we've handled all of the data that is non-numeric in the `Water_cover` column, we can change the column type to numeric. 

```{r}
# set water_cover to numeric
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Water_cover = as.numeric(Water_cover))
```

Now let's make sure that all of the values are less than or equal to 100 since these values should represent a percentage.

```{r}
# view which columns have values outside of 0 to 100
snowsurvey_fixed |> 
  filter(Water_cover > 100 | Water_cover < 0)
```

Only one value is less than 0 or greater than 100. Since the value for `Water_cover` is so much larger than 100 and `Land_cover` is negative this row of observations is unusable. For this reason we'll drop this row. 

### Drop the row with unusable values
```{r}
# keep any rows where water cover is between 0 and 100 as well as NAs
snowsurvey_fixed <- snowsurvey_fixed |> 
  filter(Water_cover <= 100 & Water_cover >= 0 | is.na(Water_cover))
```

Let's double check that worked.

```{r}
# view if columns have values outside of 0 to 100
snowsurvey_fixed |> 
  filter(Water_cover > 100 | Water_cover < 0)
```

Yay! The `Water_cover` column is clean. 

## 2. Clean the `Land_cover` column to transform it into the correct data type and respect expectations for a percentage

Let's inspect the `Land_cover` column more closely.
```{r}
# look at all of the unique values stored in the land_cover column
unique(snowsurvey_fixed$Land_cover)
```
Again, there are many non-numeric entries in this data frame (unk, n/a, ., -). We need to make them all `NA`.
```{r}
# convert any non-numeric entries to NA 
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Land_cover = ifelse(Land_cover == ".", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "-", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "unk", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "n/a", NA, Land_cover))
```

Let's make sure that no more non-numeric values other than NA are found in the `Land_cover` column. 
```{r}
# view all non-numeric values
snowsurvey_fixed  |> 
  count(Land_cover) |> 
  filter(is.na(as.numeric(Land_cover)))
```
Great! All of the `NA`'s are standardized. 

### Mutate `Land_cover` to be numeric

Now that we've handled all of the data that is non-numeric in the `Land_cover` column we can change the column type to numeric. 

```{r}
# change the class of land cover to numeric 
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Land_cover = as.numeric(Land_cover))
```

Finally, let's make sure the assumptions of percentages aren't violated. We'll make sure that the values aren't below zero and don't exceed 100.

```{r}
# view data that is outside of the 0 to 100 range
snowsurvey_fixed |> 
  filter(Land_cover > 100 | Land_cover < 0)
```

We have one row that has a value below zero. It's negative 100. Since `Snow_cover` has a value of 200 for this row, the best thing to do is drop the row. If only `Land_cover` had a value other than 0, then we would assume that there was a typo and that the value should simply be 100 instead of -100. 

```{r}
# filter out any rows that have values outside of 0 to 100 or NA
snowsurvey_fixed <- snowsurvey_fixed |> 
  filter(Land_cover <= 100 & Land_cover >= 0 | is.na(Land_cover))
```

Let's double check that there are no more rows with a value less than zero or greater than 100.
```{r}
# verify no more rows are outside of 0 to 100
snowsurvey_fixed |> 
  filter(Land_cover > 100 | Land_cover < 0)
```

Finally, before using these rows for computations, let's make sure they're all in the format we need.

```{r}
# view dataframe attributes
glimpse(snowsurvey_fixed)
```
All of the data for cover is numeric except for the `Total_cover` which we'll overwrite. 

## 3. Use the relationship that the three cover columns (Snow, Water, Land) should sum up to 100% to infer missing values where possible and recompute the `Total_cover` column as needed.

```{r}
# add snow, water, and land cover to calculate total cover
snowsurvey_fixed <- snowsurvey_fixed |> 
  rowwise() |> 
  mutate(Total_cover = sum(Snow_cover, Water_cover, Land_cover, na.rm = TRUE))
```

We also want to make sure that the `Total_cover` column is converted to numeric.
```{r}
# check the data type of the total_cover column
class(snowsurvey_fixed$Total_cover)
```


Let's look at any values that are not equal to 100.
```{r}
# save the values outside of the percentage range 
false_values <- snowsurvey_fixed |> 
  filter(Total_cover != 100)

# view the differnt values outside of the percentage range
unique(false_values$Total_cover)
```

Though in an ideal world we would salvage some of the data that is outside of the percentage range, we don't want to change the values and fabricate the information. It would take a lot of deliberation to decide what `cover` category the added or subtracted values should stem from to make the `Total_cover` equal to 100. As someone less familiar with this project, the safest route is to omit any data that does not follow the specifications of the metadata. Let's drop the rows where `Total_cover` is not equal to one-hundred as they're unusable. 

```{r}
# drop any rows where total_cover is not equal to 100
snowsurvey_fixed <- snowsurvey_fixed |> 
  filter(Total_cover == 100)
```

Save the final file to a CSV for future use.
```{r}
# write final clean file
write_csv(snowsurvey_fixed, file.path(datadir_processed, "all_cover_fixed_SOFIARODAS.csv"))
```

## Double check our new file is what we expect
```{r}
# read in the new file
clean <- read_csv(here::here("data", "processed", "all_cover_fixed_SOFIARODAS.csv"))
```

```{r}
# look at the column types
spec(clean)
```

All of the cover columns that should be numeric are numeric. 
