---
title: "eds213_data_cleaning_assign_SOFIARODAS"
format: html
author: Sofia Rodas
---
# Data Cleaning

Load libraries and data.
```{r}
#| message: false  

# load libraries
library(tidyverse)

# create a file path
datadir_raw <- file.path("data","raw")
datadir_processed <- file.path("data", "processed/")

# create variables with file names
species_file <- "ASDN_Daily_species.csv"
snowsurvey_file <- "ASDN_Snow_survey.csv"

# import the files
snowsurvey_csv <- read_csv(file.path(datadir_raw, snowsurvey_file))
```

#### View the dataframe
Before beginning to change any data types let's inspect the data frame.
```{r}
spec(snowsurvey_csv)
```
Only year is currently numeric. The remainder of the columns are characters.


## 1. Clean the Water_cover column to transform it into the correct data type and respect expectations for a percentage

Let's inspect the `Water_cover` column more closely.
```{r}
# look at all of the unique values stored in the water_cover column
unique(snowsurvey_csv$Water_cover)
```
There are many ways that `NA` is conveyed in this data frame (unk, n/a, ., -). We need to make them all `NA`.
```{r}
snowsurvey_fixed <- snowsurvey_csv |> 
  mutate(Water_cover = ifelse(Water_cover == ".", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "unk", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover))
```

Let's make sure that no more non-numeric values are found in this column. 
```{r}
snowsurvey_fixed  |> 
  count(Water_cover) |> 
  filter(is.na(as.numeric(Water_cover)))
```
Amazing! All of our data is NA when it's not numeric. 

### Mutate `Water_cover` to be numeric

Now that we've handled all of the data that is non-numeric in the `Water_cover` column we can change the column type to numeric. 

```{r}
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Water_cover = as.numeric(Water_cover))
```




### 2. Clean the Land_cover column to transform it into the correct data type and respect expectations for a percentage

### 3. Use the relationship that the three cover columns (Snow, Water, Land) should sum up to 100% to infer missing values where possible and recompute the Total_cover column as needed