---
title: "eds213_data_cleaning_assign_SOFIARODAS"
format: html
author: Sofia Rodas
---
# Data Cleaning

Load libraries and data.
```{r}
#| message: false  

# load libraries
library(tidyverse)

# create a file path
datadir_processed <- file.path("data", "processed/")

# create a variable with file names
snowsurvey_file <- "snow_cover.csv"

# import the files
snowsurvey_csv <- read_csv(file.path(datadir_processed, snowsurvey_file))
```

#### View the dataframe
Before beginning to change any data types let's inspect the data frame.
```{r}
spec(snowsurvey_csv)
```
Only `Year` and `Snow_cover` are currently numeric. The remainder of the columns are characters.


## 1. Clean the `Water_cover` column to transform it into the correct data type and respect expectations for a percentage

Let's inspect the `Water_cover` column more closely.
```{r}
# look at all of the unique values stored in the water_cover column
unique(snowsurvey_csv$Water_cover)
```
There are many non-numeric entries in this data frame (unk, n/a, ., -). We need to make them all `NA`.


```{r}
# change all non-numeric entries to NA
snowsurvey_fixed <- snowsurvey_csv |> 
  mutate(Water_cover = ifelse(Water_cover == ".", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "unk", NA, Water_cover)) |> 
  mutate(Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover))
```

Let's make sure that no more non-numeric values other than NA are found in this column. 
```{r}
snowsurvey_fixed  |> 
  count(Water_cover) |> 
  filter(is.na(as.numeric(Water_cover)))
```
Amazing! All of our data is NA when it's not numeric. 

### Mutate `Water_cover` to be numeric

Now that we've handled all of the data that is non-numeric in the `Water_cover` column we can change the column type to numeric. 

```{r}
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Water_cover = as.numeric(Water_cover))
```

Now let's make sure that all of the values are less than or equal to 100 since these values should represent a percentage.

First we'll filter for any values that are greater than 100.
```{r}
snowsurvey_fixed |> 
  filter(Water_cover > 100 | Water_cover < 0)
```

Only one value is less than 0 or greater than 100. Since the value for `Water_cover` is so much larger than 100 and `Land_cover` is negative this row of observations is unusable. For this reason we'll just drop this row. 


### Drop the row with unusable values
```{r}
snowsurvey_fixed <- snowsurvey_fixed |> 
  filter(Water_cover <= 100 & Water_cover >= 0 | is.na(Water_cover))
```

Let's double check that worked.

```{r}
snowsurvey_fixed |> 
  filter(Water_cover > 100 | Water_cover < 0)
```

Yay! The `Water_cover` column is clean. 

### 2. Clean the Land_cover column to transform it into the correct data type and respect expectations for a percentage

Let's inspect the `Land_cover` column more closely.
```{r}
# look at all of the unique values stored in the water_cover column
unique(snowsurvey_fixed$Land_cover)
```
Again, there are many non-numeric entries in this data frame (unk, n/a, ., -). We need to make them all `NA`.

```{r}
# convert any non-numeric entries to NA 
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Land_cover = ifelse(Land_cover == ".", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "-", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "unk", NA, Land_cover)) |> 
  mutate(Land_cover = ifelse(Land_cover == "n/a", NA, Land_cover))
```

Let's make sure that no more non-numeric values other than NA are found in the `Land_cover` column. 
```{r}
snowsurvey_fixed  |> 
  count(Land_cover) |> 
  filter(is.na(as.numeric(Land_cover)))
```
Great! All of the `NA`'s are standardized. 

### Mutate `Land_cover` to be numeric

Now that we've handled all of the data that is non-numeric in the `Land_cover` column we can change the column type to numeric. 

```{r}
snowsurvey_fixed <- snowsurvey_fixed |> 
  mutate(Land_cover = as.numeric(Land_cover))
```

Finally, let's make sure the assumptions of percentages aren't violated. We'll make sure that the values don't exceed 100.

```{r}
snowsurvey_fixed |> 
  filter(Land_cover > 100 | Land_cover < 0)
```

We have one row that has a value below zero. It's negative 100. Since `Snow_cover` has a value of 200 for this row, the best thing to do is drop the row. If only `Land_cover` had a value, then we would assume that there was a typo and that the value should simply be 100 instead of -100. 

```{r}
snowsurvey_fixed <- snowsurvey_fixed |> 
  filter(Land_cover <= 100 & Land_cover >= 0 | is.na(Land_cover))
```

Let's double check that there are no more rows with a value less than zero or greater than 100.
```{r}
snowsurvey_fixed |> 
  filter(Land_cover > 100 | Land_cover < 0)
```

Finally, before using these rows for computations lets make sure they're all in the format we need.

```{r}
glimpse(snowsurvey_fixed)
```
All of the data for cover is numeric except for the `Total_cover` which we'll overwrite. 

### 3. Use the relationship that the three cover columns (Snow, Water, Land) should sum up to 100% to infer missing values where possible and recompute the `Total_cover` column as needed.

```{r}
# Add snow, water, and land cover to calculate total cover
snowsurvey_fixed <- snowsurvey_fixed |> 
  rowwise() |> 
  mutate(Total_cover = sum(Snow_cover, Water_cover, Land_cover, na.rm = TRUE))
```

We also want to make sure that the `Total_cover` column is converted to numeric.
```{r}
# check the data type of the total_cover column
class(snowsurvey_fixed$Total_cover)
```


Let's look at any values that are not equal to 100.
```{r}
# save the values outside of the percentage range to 
false_values <- snowsurvey_fixed |> 
  filter(Total_cover != 100)

unique(false_values$Total_cover)
```

```{r}
false_values |> 
  filter(Total_cover <= 110 & Total_cover >= 90)
```

There are 637 observations that are between 90 and 110. These could all just be errors inputting. Since the data is collected in the Arctic 
